---
title: "Managing checkpoint snapshot archives"
author: "Andrie de Vries"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Using checkpoint for reproducible research}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `checkpoint()` function enables reproducible research by managing your R package versions. These pacakges are downloaded into a local `.checkpoint` folder.

If you use `checkpoint()` for many projects, these local packages can consume some storage space, so the package also exposes functions to manage your snapshots.

In summary:

* Managing local archives:

     * `checkpointArchives()`: list checkpoint archives on disk.
     * `checkpointRemove()`: remove checkpoint archive from disk.
     * `getAccessDate()`: returns the date the snapshot was last accessed.
     
* Other:
     * `unCheckpoint()`: reset `.libPaths` to the user library to undo the effect of `checkpoint()` (experimental).


## Setting up an example project:

For illustration, set up a script referencing a single package:

```r
library(MASS)
hist(islands)
truehist(islands)
```


Next, create the checkpoint:


```{r setup-1, include=FALSE}
## Write dummy code file to project
example_code <- '
library(MASS)
'
dir.create(tempdir(), recursive = TRUE)
cat(example_code, file = file.path(tempdir(), "managing_checkpoint_example_code.R"))
```

```{r checkpoint, results="hide", message=FALSE, warning=FALSE}
dir.create(file.path(tempdir(), ".checkpoint"), recursive = TRUE, showWarnings = FALSE)
## Create a checkpoint by specifying a snapshot date
library(checkpoint)
checkpoint("2015-04-26", project = tempdir(), checkpointLocation = tempdir())

```

## Working with checkpoint archive snapshots

You can query the available snapshots on disk using the `checkpointArchives()` function. This returns a vector of snapshot folders. 


```{r archives-1}
# List checkpoint archives on disk.
checkpointArchives(tempdir())
```

You can get the full paths by including the argument `full.names=TRUE`:

```{r archives-2}
checkpointArchives(tempdir(), full.names = TRUE)
```


## Working with access dates

Every time you use `checkpoint()` the function places a small marker in the snapshot archive with the access date. In this way you can track when was the last time you actually used the snapshot archive.

```{r access}
# Returns the date the snapshot was last accessed.
getAccessDate(tempdir())

```


## Removing a snapshot from local disk

Since the date of last access is tracked, you can use this to manage your checkpoint archives.

The function `checkpointRemove()` will delete archives from disk. You can use this function in multiple ways. For example, specify a specific archive to remove:

```{r remove-1, eval=FALSE}
# Remove singe checkpoint archive from disk.
checkpointRemove("2015-04-26")
```

You can also remove a range of snapshot archives older (or more recent) than a snapshot date


```{r remove-2, eval=FALSE}
# Remove range of checkpoint archives from disk.
checkpointRemove("2015-04-26", allSinceSnapshot = TRUE)
checkpointRemove("2015-04-26", allUntilSnapshot =  = TRUE)

```

Finally, you can remove all snapshot archives that have not been accessed since a given date:

```{r remove-3, eval=FALSE}
# Remove snapshot archives that have not been used recently
checkpointRemove("2015-04-26", notUsedSince = TRUE)

```


## Reading the checkpoint log file

One of the side effects of `checkpoint()` is to create a log file that contains information about packages that get downloaded, as well as the download size.

This file is stored in the checkpoint root folder, and is a csv file with column names, so you can read this with your favourite R function or other tools.

```{r logfile-1}
dir(file.path(tempdir(), ".checkpoint"))
```

Inspect the log file:

```{r logfile-2}

log_file <- file.path(tempdir(), ".checkpoint", "checkpoint_log.csv")
log <- read.csv(log_file)
head(log)
```


## Resetting the checkpoint

In older versions of `checkpoint()` the only way to reset the effect of `checkpoint()` was to restart your R session.

In v0.3.20 and above, you can use the function `unCheckpoint()`. This will reset you `.libPaths` to the user folder.

```{r uncheckpoint-1}
.libPaths()
```

Now use `unCheckpoint()` to reset your library paths

```{r uncheckpoint-2}
# Note this is still experimental
unCheckpoint()
.libPaths()
```


```{r cleanup, include=FALSE}
## cleanup
unlink("manifest.R")
unlink(file.path(tempdir(), "managing_checkpoint_example_code.R"))
unlink(file.path(tempdir(), ".checkpoint"), recursive = TRUE)
```
