---
title: "Using checkpoint in a markdown document"
author: "Andrie de Vries"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Using checkpoint in a markdown document}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Sometimes you want to use create a report using `markdown`, and you want to `checkpoint` the code in this document.

However, running this as part of a `RMarkdown` process is problematic, since the knitting process runs inside a temporary folder that is different from the current working directory.

To resolve this, I propose a hacky solution: create a "manifest" file in the same folder that contains all of the `library()` calls.

# Example

Imagine you have a small script that you want to put in an Rmarkdown document with a checkpoint.

```r
# demo script
library(MASS)
hist(islands)
```

# The checkpoint solution

The only way I've found to get `checkpoint()` to work inside an RMardown document, is really a bit of a hack. The workaround is to create a manifest of required packages, and write this to an R file in the working directory.

For example, imagine your R script uses the `MASS` package, then create a manifest file like this:

```{r, echo=FALSE, comment=""}
cat("```{r, include=FALSE}", '
# write a manifest to local folder
cat("
library(MASS)
",
file = "manifest.R")
', "```", sep = "")
```

This is hacky, since it requires you to construct the list of `library()` calls by hand and put these into the manifest file.

(Note that you can use `include=FALSE` in the code block, so that this code doesn't show up in your rendered document.)

Once this is done, the checkpoint process from here is straight-forward


```{r, echo=FALSE, comment=""}
cat("```{r, include=FALSE}", '
# Create .checkpoint folder (in tempdir for this example)
td <- tempdir()
dir.create(file.path(td, ".checkpoint"), recursive = TRUE)

# Create the checkpoint
library(checkpoint)
checkpoint("2017-03-28", checkpointLocation = td)
', "```", sep = "")
```

# Check that this works

Now you are ready to put these instructions in an actual code block to see what happens.

```{r checkpoint, warning=FALSE}
# write a manifest to local folder
cat('
library(MASS)
',
file = "manifest.R")

# Create .checkpoint folder (in tempdir for this example)
dir.create(file.path(tempdir(), ".checkpoint"), recursive = TRUE)

# Create the checkpoint
library(checkpoint)
checkpoint("2017-03-28", checkpointLocation = tempdir())

```


If this worked, you should see that the library path now points to `tempdir()` and that MASS should be one of only a few package installed:

```{r, eval=FALSE}
.libPaths()
## [1] ".../Temp/RtmpIVB6bI/.checkpoint/2017-03-28/lib/x86_64-w64-mingw32/3.3.2"
## [2] ".../Temp/RtmpIVB6bI/.checkpoint/R-3.3.2"
```

```{r}
installed.packages()[, "Package"]
```

# Your real R code:

Now your real R code follows, and it creates the plot, as expected:

```{r, warning=FALSE}
library(MASS)
hist(islands)
```

# Conclusion

This is a bit of a hack, but points in a direction for getting your RMarkdown script to be checkpointed.


```{r, include=FALSE}
# clean up

detach("package:MASS", unload = TRUE)
unlink("manifest.R")
unlink(file.path(tempdir(), ".checkpoint"), recursive = TRUE)
unCheckpoint()
```
